### 2022-09-03 chapters/optimization/python-dir/python_interperters.tex
fix : notes that were hilighlted by Lionel
update:  python - part 1
python interpreters

%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs
% TODO : ADD DRAM graphs for vectors and matrix multiply

@@

\section{Results and finding}
This section will be dedicated to the results of the experiments and the finding of the best implementation.
\paragraph{python and binarry operations }
The goal of this study is to look at how different Python implementations work while treating different binary operations .

\begin{figure}[!hbt]
    \begin{tabular}{|lrrrrr|c|}
        \toprule
        benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average \\
        \midrule
        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
        cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
        \hline
        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
        \hline
        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
        \bottomrule
    \end{tabular}
    % should i add the interpretation 
    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
    \label{table:bitops}
\end{figure}

@@

Therefore, it is safe to summarize all the operations into one that we call binary.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
    \caption{Summary of the binary operations for different python implementations }
    \label{figure:bitops}
\end{figure}
Figure \ref{figure:bitops}  shows the average energy consumption of different implementations for the binary operations.

One can notice 3 clusters.

@@

\section{introduction to the study}
Since we are in a comparative study, we first performed a shapiro normality test on the results, and almost all the p-values are  smaller than alpha=0.01\% therefore the distrubution is not normal. To make a comparaison we used Manwithyu's tests.
Table \ref{table:manwithyu} .

\input{table-envs-based-cpython.tex}
for arrays and vectors  graalpython took so much time that we decided to remove it

the startup cost for ironpython was too high
for graalpython and ipy they couldnt handle big numbers hence the overflow  when it comes to execution time
but since we measure the energy outside the values of the energy aren't impacted

as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU

then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage. howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read

\input{table-envs.tex}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
    \caption{Dendogram of the different implementations}
    \label{fig:dendogram}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{imgs/alltomti_performance}
    \caption{different interpererts optimisation }
    \label{fig:tommi_all}
\end{figure}

%   this is after , when we come to the conclusion
As we can see in figure \ref{fig:tommi_all}, there is no evolution between cpython2 and cpython3
Intelpython and active python both folows the same behaviour. one can conclude that the work that have been on those interpreters is mainly
to improve a specific purpose, Active python claims that their version is focused on secuirty which explains the lack of some performances due to the introduction of more reonforcment
.Morver Intel published their version of python as a dedicated for machine learning. Unfortunately the tommti benchmark is a set that focuses on the general purpose programming
which does not reflect the performance of the machine learning.
Another aspect of a such behaviour might be due to the fact of the processors that were used in the testbed. and it might change in the future if we include the GPU part for the machine learning benchmarks
% reference the work of the intern
As for nuitka. there were no optimization in the energy consumpotion despite the fact that is it a complier.
However if we dig through the nuitka mechanisms. they basically embed the python code with an interpreter.
Unllike nuitka shedskin exhebit a the best energy consumption pattern when it comes to the arithmetic operartions. One can conclude it is due to the fact of the native type of the variables. unlike the interpreters where they are treated as object in the begining.

for the other interpeters pypy is very promising especially when it comes to data maniupulation as one can see in the figure \ref{fig:bar_tommtei_vector}
pypy is by far the best interperer when it comes to treating vectors.
numba2 introduced the JIT but wasn't as promising as numba3.

for the other vm based interpereters. jython and ipy lacked in term of energy optimisation which was kinda expected since they were in their the begining of the stage and the main puprose of such implementation is to link the bytecode generated by jython and ironpython with their respectives virtual machin.

Unlike the the previous interpeters. graal exhebits a certin promises when it comes to complex algorithms - nested loops -
micro python is dedicated to embeded systems so lunching it is powerful cpu machines will be misleading.

Most of the interperters had the same behaviour when it comes to the input outputs. except for jython which was kinda of a anomaly probably due to the lack of the optimizations.
% pypy handles exceptions very well

\section{conclusion}
One may observe that the choice of Python interpreter has a significant impact on the programs' energy consumption.
This investigation is made more intriguing by the absence of a universal solution.
The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.

>>>>

%% basically i what i want to say is that we we care mostly about energy, because there are studies that were made on the performance, plus the DRAM is not that interested even within the heavy memory  benchmakrs

@@

\section{Results and finding}

This section will be dedicated to the results of the experiments and a statistical analysis of these results.

Because we are conducting a comparison study, we first did a Shapiro-Wilk~\cite{shapiro1968comparative} test to see if the data follows a normal distribution.
Despite, the fact some implementations had a p-value bigger than a=0.05 (such as cpython and active python),other implementaions like pypy and numba presented a p-values smaller than 0.01 which leads to reject the hypothesis H0 that all the distributions follow a normal distribution. 
Therefore,to compare the different implementations, we performed the non-parametric Mann-Whitney U test~\cite{zimmerman1987comparative}, with CPython2 and CPython3 as base references.

The Mann-Whitney U test results for each implementation are shown in Table \ref{table:manwithyu}. The first column displays the implementation's average energy usage, while the second provides the p-value of the test when compared to the reference Python implementation. If the p-value is less than 0.05, the test result is significant. This implies that the given implementation's energy consumption differs significantly from the reference implementation's.As one can notice, most of the implemnetations differe significantly from the reference implementation. 
\input{table-envs-based-cpython.tex}
We want to study the position of these implementations toward the references. To do so, we proceed with an agglomerative hierarchical cluster analysis (HCA)~\cite{johnson1967hierarchical}.
The HCA is a method that allows grouping the different implementations based on their energy consumption. This technique is a bottom-up approach that starts with each implementation as a single cluster and then merges the two most similar clusters until all the clusters are merged into a single cluster. The similarity between two clusters is measured by the maximum distance between their furthest points. The distance between two points is calculated as the Euclidean distance between the two points.
Figure \ref{fig:dendogram} displays the dendogram produced by the HCA
\begin{figure}
      \centering
      \includegraphics[width=\linewidth]{imgs/dendogram_interpreters}
      \caption{Dendogram of the different implementations}
      \label{fig:dendogram}
\end{figure}

Except GraalPython, one can notice 3 main clusters.
The first cluster contains the reference implementations (CPython2 and CPython3) and the implementations that are based on interpreters (Ironpython,Intelpython, ActivePython, Jython, and Nuitka). Moreover, each implementation is the closest to its reference Python version. The interpreters that are based on other virtual machines, such as Jython with JVM and IronPython with.NET, behave slightly differently from the reference implementation. As for the micropython, It is the furthest from the references due to his behavior toward exceptions.
The second cluster contains the interpreter implementations that are based on a JIT compiler (Pypy2 and Pypy3). Closer to them we find shedskin, which is a transaltor that converts the Python code into C++ code to be later compiled into binary code.
The results table~\ref{table:manwithyu} shows that this cluster is by far the greenest one.
While numba2 and numba3 are also JIT libraries, they differ from PyPy due to their manual optimisation. Unlike pypy, which will decide by itself which part of Python to optimize. In Numba, the developer must target the portion of the code that will be passed to JIT, while the rest of the code will be executed by the regural interpreter (aka CPython).

\paragraph{Remark :}
Because graalPython was still in its early stages, some techniques required an abnormally lengthy time, affecting the clustering Algorithm.

To help practiouners to choose the best implementation for their use case, we have created a chart that describes the behavior of each implementation toward a specific aspect of programming. 
Figure~\ref{fig:tommi_all} present a radar plot for all the implementations that have been used in this experemation.
\begin{figure}[!htb]
      \centering
      \includegraphics[width=\linewidth]{imgs/alltomti_performance}
      \caption{Different interpererts optimisation score }
      \label{fig:tommi_all}
\end{figure}

This graph depicts each implementation's energy score in relation to each benchmark. The lower the energy consumption, the higher the score. We adopted a logarithmic scale to allow the reader to compare numerous implementations due to the large differences between them.

As one can notice in figure~\ref{fig:tommi_all}, there is no evolution between CPython2 and CPython3. Moreover, Intelpython, and Active Python all behave similarly to the reference implementation . Therefore, one can conclude that the work done on those interpreters is primarily to improve a specific purpose and not the core interpreter.
Active Python states that their version is focused on security and prepackaged libraries, which explains why it is slower than other versions due to the addition of more security reinforcements\footnote{\url{https://www.activestate.com/solutions/why-activestate/}} .
As for IntelPython,it is designed for machine learning. Unfortunately, the Tommti benchmark is geared for general-purpose programming.
Despite the fact that Nuitka is a compiler, there was no discernible difference in energy consumption. It was even more similar to the reference Python implementation than other interpreters.
However, if we look at the nuitka techniques, we can see that they simply encapsulate the Python code with an interpreter into a single executable.
However, shedskin exhebit has the best energy consumption pattern when it comes to arithmetic operations. One can conclude it is due to the fact of the native type of the variables. Unlike in the JIT, where they are treated as objects in the beginning.

for the other VM-based interpereters. Jython and Ipy lacked in terms of energy optimisation, which was kind of expected since they were at the beginning of the development stage and the main purpose of such implementation is to link the bytecode generated by Jython and IronPython with their respective virtual machines.

Unlike the previous interpeters, GraalPython exhibits a certain promise when it comes to complex algorithms (nested loops).
Finally,micro python is dedicated to embedded systems, so claiming it is a powerful CPU machine will be misleading.

When it comes to  the input outputs operations , 
when it came to input/output manupulation. Except for Jython,most of the interpreters behaved similarly . Because these operations are more dependent on the hardware than the interpreter.

% However, Jython is the only implementation that has a significant difference in energy consumption. This is due to the fact that Jython is based on the JVM, which is a virtual machine that is designed to run Java bytecode. Therefore, the energy consumption of the input/output operations is not the same as the other implementations.

% \begin{figure}
%       \centering
%       \includegraphics[width=\linewidth]{imgs/bar_plot_tommtivector}
%       \caption{Average energy consumption of different python implementaions for vector treatement}
%       \label{fig:bar_tommti_vector}
%   \end{figure}
% as usual there is  correlaction between DRAM and CPU so no need to classify based on DRAM + the DRAM consumes less 10\% of the  energy compared to the CPU

% then we use the multiparameter optimization , howerver we stop in the phase where we calculate the score
% because we dont know the weight of each parameter ( of tommti microbenchmark) and instead of doing the sum with the coefissients we use the radar plot to let the reader decide which one is the most suitable for his usage. howerver since the differences are gigantics we change the scale to logarithmic to be easier for the eye to read

% \input{table-envs.tex}

%% distance euclidean 
%% linkage complete Voor Hees Algorithm.

%%%%%%%%%%%%%%%%%

%   this is after , when we come to the conclusion
% pypy handles exceptions very well

\subsection{python and binarry operations }
The second part of this study, is to look at how different Python implementations work while treating different binary operations .

\begin{figure}[!hbt]
    \begin{tabular}{|lrrrrr|c|}
        \toprule
        benchmark    & Addition & Right Rotation & Left Rotation & Or      & XOR     & Average \\
        \midrule
        activepython & 676.980  & 763.208        & 651.783       & 743.016 & 728.828 & 712.76  \\
        cpython2     & 441.082  & 435.886        & 430.846       & 415.247 & 419.081 & 428.43  \\
        cpython3     & 595.209  & 685.085        & 563.839       & 657.972 & 655.560 & 531.53  \\
        \hline
        cython       & 35.077   & 182.688        & 274.177       & 34.868  & 34.504  & 112.26  \\
        nuitka       & 33.260   & 32.980         & 33.256        & 33.472  & 33.030  & 33.2    \\
        \hline
        numba2       & 9.102    & 8.411          & 9.460         & 9.375   & 9.755   & 9.22    \\
        numba3       & 9.566    & 10.144         & 9.219         & 9.344   & 9.665   & 9.59    \\
        pypy2        & 8.456    & 7.844          & 8.286         & 8.138   & 7.952   & 8.13    \\
        pypy3        & 7.552    & 8.093          & 8.108         & 8.669   & 8.623   & 8.21    \\
        shedskin     & 8.024    & 8.070          & 8.399         & 8.126   & 8.277   & 8.18    \\
        \bottomrule
    \end{tabular}
    \caption{energy consumption of different impelementations using Bit Operation benchmarks (Joule) }
    \label{table:bitops}
\end{figure}

@@

Therefore, it is safe to summarize all the operations into one that we call \em{binary}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{imgs/bitopts_mean}
    \caption{Summary of the binary operations for different python implementations }
    \label{fig:bitops}
\end{figure}
Figure \ref{fig:bitops}  shows the average energy consumption of different implementations for the binary operations.One can notice 3 clusters.

@@

\section{conclusion}
One can conclude that the choice of Python interpreter has a significant impact on the programs' energy consumption.
This investigation is made more intriguing by the absence of a universal solution.
The primary downside is the incompatibility of some of these solutions, which causes us to make concessions when we need a generic answer.
